---
title: "Bayesian Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, eval = TRUE}
library(mlstats)
```

The frequentist functions `mldesc()` and `within_between_correlations()` rely on
null-hypothesis tests: exact t-tests (under normality) for the default
`method = "decomposition"`, or asymptotic z-tests for `method = "sem"` (see
`vignette("correlation-methods")` for details on both). Their Bayesian counterparts,
`bayes_mldesc()` and `bayes_within_between_correlations()`, instead report
**posterior medians** with **credible intervals** (CIs): a correlation is
flagged with a star when its CI excludes zero. The output format is otherwise
identical, so you can switch between the two approaches without changing your
downstream code.

The Bayesian functions require the **brms** package, which is not installed
with **mlstats** by default:

```{r install-brms}
install.packages("brms")
```

## When to choose Bayesian over frequentist

The frequentist decomposition method works well for most data sets, but the
Bayesian approach has advantages in specific situations:

- **Few groups.** When the number of groups is small (e.g., fewer than 30
  persons in a diary study), the significance test for the between-group
  correlation becomes harder to trust either way: the decomposition method's
  exact t-test assumes the group means are bivariate normal, an assumption
  that is difficult to verify with only a handful of groups, and `method =
  "sem"`'s asymptotic MLR standard errors can also become unreliable with few
  clusters. Bayesian posteriors do not rely on either assumption.
- **Uncertainty quantification.** CIs convey the precision of each estimate
  directly, without requiring a separate look at standard errors or effect sizes.

When group sizes are large and you have many groups, the frequentist functions
are faster and the results will be nearly identical.

## The `folder` argument and model caching

Every call to a Bayesian function requires a `folder` argument specifying where
fitted `brms` models should be saved. Models are cached on disk, so re-running
the same call with the same data and settings is instant — the saved file is
reloaded instead of refitting. A content hash of the data and sampling settings
is folded into every filename, so changing either the data or
`options(mlstats.brms_iter = ...)` automatically triggers a refit rather than
silently loading a stale result.

```{r folder-setup}
brms_dir <- "brms_models"
```

## Basic usage

The examples below use `media_diary`, the simulated daily diary dataset included
with **mlstats** (100 participants over 14 days; *N* = 100 persons, *T* =
1,400 daily observations; see `?media_diary` and `vignette("mlstats")` for
details).

```{r data}
data("media_diary")
vars <- c("self_control", "wellbeing", "screen_time", "stress")
```

`bayes_within_between_correlations()` has the same interface as
`within_between_correlations()`, plus the `folder` and `ci` arguments. `method`
and `significance` are not applicable here: estimation is always via `brms`,
and significance marking is always based on whether the credible interval
excludes zero (controlled by `ci`).

```{r basic-correlations}
bayes_within_between_correlations(
  data = media_diary,
  group = "person",
  vars = vars,
  folder = brms_dir
)
```

`bayes_mldesc()` adds descriptive statistics and ICCs:

```{r basic-mldesc}
result <- bayes_mldesc(
  data   = media_diary,
  group  = "person",
  vars   = vars,
  folder = brms_dir
)

result
```

## Credible interval width

The `ci` argument controls the width of the credible interval used to determine
whether a correlation is marked with a star. The default is `0.9` (90% CI). A
correlation is starred when the CI excludes zero.

```{r ci}
bayes_mldesc(
  data   = media_diary,
  group  = "person",
  vars   = vars,
  ci     = 0.95,
  folder = brms_dir
)
```

The ICC in the rightmost column always reports the posterior median and is
not affected by `ci`.

## Adjusting sampling settings

`brms` uses Markov chain Monte Carlo (MCMC) sampling. The defaults (5,000
iterations, 4 chains) are conservative and usually sufficient, but you can
change them globally for the session:

```{r options}
options(
  mlstats.brms_iter   = 2000,
  mlstats.brms_chains = 2
)
```

Reducing iterations speeds up exploration runs; increase them when you need
more stable posterior estimates or when `brms` reports convergence warnings
(R-hat > 1.01, low effective sample sizes).

## Number of models fitted

For `p` variables, `bayes_within_between_correlations()` fits:

- `p * (p - 1) / 2` within-group models (one per pair)
- `p * (p - 1)` between-group models when `weight = TRUE` (two per pair:
  one weighted for the point estimate, one unweighted for the CI), or
  `p * (p - 1) / 2` when `weight = FALSE`

`bayes_mldesc()` additionally fits `p` intercept-only models for the ICCs.
For 4 variables with default settings, this is 6 within-group + 12
between-group + 4 ICC = **22 models**. Plan accordingly.

## Formatted output

The result object is a `mlstats_desc_tibble` with the same print methods as
`mldesc()`:

```{r formatted-output}
# tinytable (included with mlstats)
print(result, format = "tt",
      table_title = "Bayesian multilevel descriptive statistics")

# gt table (requires the gt package)
print(result,
  format           = "gt",
  table_title      = "Bayesian multilevel descriptive statistics",
  correlation_note = "Within-person correlations above, between-person below the diagonal.",
  note_text        = "<i>Note</i>. <i>N</i> = 100, <i>T</i> = 1,400 daily observations. 90% credible intervals."
)
```

See `vignette("tables")` for a full guide to table customisation.

## Comparison with the frequentist output

The only differences between `bayes_mldesc()` and `mldesc()` output are:

| Feature | Frequentist | Bayesian |
|---|---|---|
| Significance criterion | *p* < .05 | CI excludes zero |
| CI width | — | Controlled by `ci` |
| Speed | Fast (seconds) | Slow (minutes) |
| Minimum group requirement | ~30+ for reliable between CIs | Works with fewer groups |

For large, well-powered studies with many groups, the two approaches give nearly
identical results. For small studies or when communicating uncertainty is a
priority, the Bayesian route is preferable.

## References

Bürkner, P.-C. (2017). brms: An R package for Bayesian multilevel models using
Stan. *Journal of Statistical Software, 80*(1), 1–28.
https://doi.org/10.18637/jss.v080.i01
