---
title: "Getting Started with mlstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mlstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **mlstats** package provides tools for multilevel descriptive statistics
and data preparation. It supports:

- Decomposing variables into within-group and between-group components for
  Random Effects Within-Between (REWB) models
- Computing within-group and between-group correlations
- Creating publication-ready descriptive statistics tables with ICCs
- Both frequentist (via lme4) and Bayesian (via brms) estimation

## Setup

```{r setup}
library(mlstats)
```

## Example Data

We use simulated data from 50 students nested within 10 schools, where both
school-level and student-level variation exist in math scores, reading scores,
and motivation.

```{r data}
set.seed(42)
n_schools <- 10
n_per_school <- 5

school_math <- rep(rnorm(n_schools, 50, 8), each = n_per_school)
school_read <- rep(rnorm(n_schools, 50, 8), each = n_per_school)
school_motiv <- rep(rnorm(n_schools, 5, 1.5), each = n_per_school)

data <- data.frame(
  school = rep(1:n_schools, each = n_per_school),
  math_score = school_math + rnorm(n_schools * n_per_school, 0, 5),
  reading_score = school_read + rnorm(n_schools * n_per_school, 0, 5),
  motivation = school_motiv + rnorm(n_schools * n_per_school, 0, 1)
)

head(data)
```

## Decomposing Variables

The `decompose_within_between()` function splits each variable into:

- **Grand mean centered** values (deviation from overall mean)
- **Between-group** component (group means)
- **Within-group** component (deviation from group mean)

These components are useful for fitting Random Effects Within-Between (REWB)
models (Bell et al., 2019), which separate within-group from between-group
effects.

```{r decompose}
decomposed <- decompose_within_between(
  data = data,
  group = "school",
  vars = c("math_score", "motivation")
)

# View the decomposed math_score columns
math_cols <- grep("school|math_score", names(decomposed))
head(decomposed[, math_cols], 10)
```

## Within-Group and Between-Group Correlations

The `within_between_correlations()` function computes a correlation matrix with
within-group correlations in the upper triangle and between-group correlations
in the lower triangle:

```{r correlations}
vars <- c("math_score", "reading_score", "motivation")

within_between_correlations(
  data = data,
  group = "school",
  vars = vars
)
```

### SEM-Based Estimation

By default, correlations are estimated via variance decomposition
(`method = "decomposition"`), following Pedhazur (1997). Alternatively, you
can use a two-level structural equation model via `lavaan`:

```{r correlations-sem}
within_between_correlations(
  data = data,
  group = "school",
  vars = vars,
  method = "sem"
)
```

The SEM method automatically detects between-only variables (those with zero
within-cluster variance, e.g., time-invariant traits) and within-only variables
(ICC ≈ 0). Correlations that cannot be estimated at a given level are reported
as `NA`.

## Multilevel Descriptive Statistics

The `mldesc()` function combines descriptive statistics, correlations, and ICCs
into a single publication-ready table:

```{r mldesc, warning = FALSE}
result <- mldesc(
  data = data,
  group = "school",
  vars = vars
)

result
```

### Customization Options

Control weighting, significance marking, and formatting:

```{r mldesc-options, warning = FALSE}
mldesc(
  data = data,
  group = "school",
  vars = vars,
  weight = FALSE,
  significance = "detailed",
  flip = TRUE,
  remove_leading_zero = FALSE
)
```

### Table Output Formats

The output can be rendered as a `tinytable` table for use in documents:

```{r tt-output, warning = FALSE}
result |>
  print(format = "tt")
```

For `gt` table output (requires the **gt** package), use
`print(result, format = "gt")`. See the
[package README](https://github.com/felixdidi/mlstats) for detailed examples
of formatted table output.

### Pipe-Friendly Output

The output tibbles contain vectors of class `mlstats_stat` (abbreviated `mls`).
These can be cast to numeric for subsequent calculations:

```{r pipe-friendly}
# Extract correlations
cors <- within_between_correlations(
  data = data,
  group = "school",
  vars = vars
)

# Convert to numeric (removes significance stars)
cors_numeric <- cors
cors_numeric[-1] <- lapply(cors[-1], as.numeric)
cors_numeric
```

## Bayesian Estimation

For Bayesian estimation with credible intervals instead of *p*-values, use
`bayes_mldesc()` and `bayes_within_between_correlations()`. These functions
require the **brms** package and may take considerable time to run. See
`?bayes_mldesc` for details.

## References

Bell, A., Fairbrother, M., & Jones, K. (2019). Fixed and random effects
models: making an informed choice. *Quality & Quantity*, 53(2), 1051--1074.

Enders, C. K., & Tofighi, D. (2007). Centering predictor variables in
cross-sectional multilevel models: A new look at an old issue.
*Psychological Methods*, 12(2), 121--138.
