---
title: "Publication-Ready Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publication-Ready Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(mlstats)
library(dplyr)
library(stringr)
```

`mldesc()` and `bayes_mldesc()` return a tibble that can be printed in three
formats: a console-friendly default, a **tinytable** object, and a **gt**
object. This vignette shows how to move from the default output to a
fully-customised, journal-ready table.

## Example data

We use `media_diary`, a simulated daily diary dataset included with **mlstats**
(100 participants over 14 days; *N* = 100 persons, *T* = 1,400 daily observations). See `?media_diary` for details.

```{r data}
data("media_diary")
vars <- c("self_control", "wellbeing", "screen_time", "stress", "enjoyment")
```

```{r compute, warning = FALSE}
result <- mldesc(data = media_diary, group = "person", vars = vars)
```

## Default console output

Simply printing the result gives a compact console-friendly view:

```{r console}
result
```

## tinytable format

`tinytable` is a lightweight table package included with **mlstats** (no extra
installation needed). Pass `format = "tt"` to `print()`:

```{r tt, warning = FALSE}
print(result, format = "tt")
```

The result is a `tinytable` object that renders to HTML, PDF, or Word via Quarto/R Markdown (see below).

### Custom title and notes

All print methods accept `table_title`, `correlation_note`,
`significance_note`, and `note_text`:

```{r tt-custom, warning = FALSE}
print(result,
  format           = "tt",
  table_title      = "Daily diary study: descriptive statistics and multilevel correlations",
  correlation_note = "Within-person correlations above, between-person below the diagonal.",
  note_text        = "N = 100 persons, 14 daily observations each. Simulated data."
)
```

## gt format

`gt` produces richly formatted HTML tables and supports markdown in cells,
footnotes, and fine typographic control. It must be installed separately:

```{r gt-install, eval = FALSE}
install.packages("gt")
```

```{r gt-basic}
print(result, format = "gt")
```

`gt` tables support further customisation via the `gt` package API after the
initial `print()` call — see the [gt documentation](https://gt.rstudio.com/)
for details.

## Manipulating the result before printing

Because `mldesc()` returns a tibble, standard `dplyr` operations work on it
before printing.

### Removing columns

Drop columns you don't need in the final table:

```{r drop-cols, warning = FALSE}
result |>
  select(-n_obs, -range) |>
  print(format = "tt")
```

### Replacing NA with a dash

`self_control` is a between-person-only trait: its within-person correlations
are `NA`. Replace these with an em dash for cleaner output:

```{r replace-na, warning = FALSE}
result |>
  mutate(across(everything(), ~ str_replace(as.character(.x), "^NA$", "–"))) |>
  print(format = "tt")
```

### Renaming variables

Variable names are auto-formatted as sentence case. To customise them:

```{r rename, warning = FALSE}
result |>
  mutate(variable = case_when(
    variable == "Self control" ~ "Trait self-control",
    variable == "Wellbeing"    ~ "Daily wellbeing",
    variable == "Screen time"  ~ "Screen time (min)",
    variable == "Stress"       ~ "Perceived stress",
    variable == "Enjoyment"    ~ "Media enjoyment"
  )) |>
  print(format = "tt", table_title = "Study variables: descriptive statistics")
```

### Combining manipulations

All of the above can be chained. Here is an example of a polished table
combining several customisations:

```{r combined, warning = FALSE}
result |>
  select(-n_obs, -range) |>
  mutate(across(everything(), ~ str_replace(as.character(.x), "^NA$", "–"))) |>
  mutate(variable = case_when(
    variable == "Self control" ~ "Trait self-control",
    variable == "Wellbeing"    ~ "Daily wellbeing",
    variable == "Screen time"  ~ "Screen time (min)",
    variable == "Stress"       ~ "Perceived stress",
    variable == "Enjoyment"    ~ "Media enjoyment"
  )) |>
  print(
    format           = "tt",
    table_title      = "Descriptive statistics and multilevel correlations",
    correlation_note = "Within-person correlations above, between-person below the diagonal.",
    note_text        = "N = 100, T = 1,400 daily observations. Self-control was measured as a trait (between-person only)."
  )
```

For the equivalent using `gt` (which additionally supports footnotes and
markdown-formatted cell content):

```{r combined-gt}
result |>
  select(-n_obs, -range) |>
  mutate(across(everything(), ~ str_replace(as.character(.x), "^NA$", "–"))) |>
  mutate(
    variable = case_when(
      variable == "Self control" ~ "Trait self-control<sup>c</sup>",
      variable == "Wellbeing"    ~ "Daily wellbeing",
      variable == "Screen time"  ~ "Screen time (min)",
      variable == "Stress"       ~ "Perceived stress",
      variable == "Enjoyment"    ~ "Media enjoyment"
    )
  ) |>
  print(
    format           = "gt",
    table_title      = "Descriptive statistics and multilevel correlations",
    correlation_note = "Within-person correlations above, between-person below the diagonal.",
    note_text        = "<i>Note</i>. <i>N</i> = 100, <i>T</i> = 1,400 daily observations."
  ) |>
  gt::tab_source_note(
    source_note = gt::html(
      "<sup>c</sup> Self-control was measured as a stable trait; no within-person correlations are available."
    )
  ) |>
  gt::fmt_markdown(columns = variable)
```

## Embedding in Quarto documents

### Word / DOCX output

Wrap the `print()` call in a Quarto code chunk with `format: docx`:

````{verbatim, eval = FALSE}
---
format: docx
---

```{r}
library(mlstats)
data("media_diary")

mldesc(
  data  = media_diary,
  group = "person",
  vars  = c("self_control", "wellbeing", "screen_time", "stress")
) |>
  print(format = "tt")
```
````

`tinytable` automatically converts to the appropriate format based on the output
Quarto is rendering to.

### HTML / PDF

Both `tinytable` and `gt` render natively to HTML and LaTeX. No extra setup is
needed:

````{verbatim, eval = FALSE}
---
format: html   # or pdf
---

```{r}
library(mlstats)
data("media_diary")

mldesc(
  data  = media_diary,
  group = "person",
  vars  = c("self_control", "wellbeing", "screen_time", "stress")
) |>
  print(format = "tt")
```
````
